Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-06-14 submission/manuscript.Rmd
add intro fixes suggested by Nikc

Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US [@Haggar2009; @Siegel2016]. Although there is a genetic component to the disease, the environment has been attributed as an important and large risk factor for CRC [@Lichtenstein2000]. This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered [@Shah2017; @meta_analysis_crc_Sze2018]. Multiple reports in case/control studies on the microbiota and CRC have identified taxa commonly associated with the normal colons to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @Kostic2011]. Many of these taxa actively produce short-chain fatty acids (SCFAs) from fiber that are a part of our general diet [@Holscher2017]. The most extensively studied of these SCFAs are acetate, butyrate, and propionate [@Louis2016]. Overall, CRC is a growing health problem and the microbiota through SCFAs have shown promise as a potential treatment option.

@@

There is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations. The initial case/control studies that investigated SCFA concentrations in CRC found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus either patients with adenomas or individuals with normal colons [@Ohigashi2013]. Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [@Yao2017]. These findings would argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis. Given the lack of clear evidence in human studies of the benefit of SCFAs in CRC, there is a need for more investigation into this area.

Our current study will fill some of the current gaps in the literature that relate to the study of SCFAs and CRC in human populations. Specifically, it will attempt to replicate previous case/control findings on SCFA concentrations in individuals with and without tumors. Additionally, by increasing the adenoma tumor cohort it we will identify if the observation of a continuous gradual reduction in SCFA concentrations occurs as tumor severity increases. It also will build upon these observations and assess the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a potential risk stratification tool of colorectal tumors (adenoma or carcinoma). This study will provide important information on the replicability of previous findings in humans by extensively studying how SCFAs could be associated with colorectal tumors.

To accomplish this task we will directly measure the concentration of acetate, butyrate, and propionate within fecal samples for two different groups. The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors. Additionally, we also assessed whether adding SCFA concentrations to OTU data improves the classification of adenoma or carcinoma using the Random Forest algorithm [@randomforest_citation_2002], used PICRUSt [@Langille2013] and metagenomic sequencing to assess the genes involved in SCFA synthesis, and analyzed how well 16S rRNA gene sequencing can predict SCFA concentrations. This  investigation provides additional information as to whether the taxa that are decreased in carcinoma use SCFAs to help protect against colon tumorigenesis.

>>>>

Colorectal cancer (CRC) is currently the third highest cancer-related cause of death within the US [@Haggar2009; @Siegel2016]. Although there is a genetic component to the disease, the environment is attributed to being a larger risk factor for CRC [@Lichtenstein2000]. These environmental risk factors include but are not limited to smoking cigarettes, diet, and the microbiota [@FlissIsakov2017; @Lee2015; @Kostic2011]. Many of these environmental risk factors are capable of being modified, and this has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered [@Shah2017; @meta_analysis_crc_Sze2018]. Multiple reports in case/control studies have identified bacterial taxa commonly associated with individuals without tumors to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @meta_analysis_crc_Sze2018]. Many of these taxa within individuals wihtout tumors actively produce short-chain fatty acids (SCFAs) from fiber that are a part of our general diet [@Holscher2017]. The most extensively studied of these SCFAs are acetate, butyrate, and propionate [@Louis2016]. Overall, the specific bacterial taxa of the microbiota that create SCFAs are an attractive target to modulate the risk of CRC.

@@

There is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations. The initial case/control studies that investigated SCFA concentrations in CRC found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus either patients with adenomas or individuals without colon tumors [@Ohigashi2013]. Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [@Yao2017]. These findings would argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis. Given the lack of clear evidence in human studies of the benefit of SCFAs in CRC, there is a need for more investigation into this area.

Our study fills some of the current gaps in the literature that relate to the study of SCFAs and CRC in human populations. Specifically, it tests previous case/control findings on SCFA concentrations in individuals with and without tumors. We also test previous suggestions that there is a continuous reduction in SCFA concentrations as tumor severity increases by increasing the number of patients with adenomas in our study. Additionally, we build upon these observations and assesses the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a risk stratification tool of colorectal tumors (adenoma or carcinoma). Collectively, this study provides important information on the replicability of previous findings in humans by extensively studying how SCFAs are associated with colorectal tumors.

To accomplish this task we directly measured the concentration of acetate, butyrate, and propionate within fecal samples for two different groups. The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors. Additionally, we (i) assessed the affect adding SCFA concentrations to OTU data had on classification of patients with adenoma or carcinoma using the Random Forest algorithm [@randomforest_citation_2002], (ii) used PICRUSt [@Langille2013] and metagenomic sequencing to assess the presence of genes involved in SCFA synthesis, and (iii) analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations. This investigation provides additional information as to whether SCFAs are decreased in patients with colorectal tumors and provides context as to whether targeting taxa to increase SCFA concentrations is a viable option to protect against colon tumorigenesis.
